### 2023-06-08 bachelorthesis/01-dedication.tex
#39: Correction Cornelia Lippert

\begin{center}
    \vspace*{3cm}
    \begin{tikzpicture}
        \clip(0,0)circle[x radius=3cm, y radius=4cm];
        \node{\includegraphics[width=6cm]{Dedication/blacky1_black_white.jpg}};
    \end{tikzpicture}
    \vspace{2cm}\\
    (\textborn) 29.05.2005 - (\textdied) 12.05.2023\\
    \bigskip
    This thesis is dedicated to my cat \textbf{Blacky}\\ which get nearly 19 years old and followed me\\ since I was 5 years through my life.\\You were always a big part of my life and\\ 
    \bigskip
    I will miss you.\\
    \bigskip
    Manuel Lippert\\
    \today
\end{center}

>>>>

\begin{center}
    \vspace*{3cm}
    \begin{tikzpicture}
        \clip(0,0)circle[x radius=3cm, y radius=4cm];
        \node{\includegraphics[width=6cm]{Dedication/blacky1_black_white.jpg}};
    \end{tikzpicture}
    \vspace{2cm}\\
    \gtrsymBorn~29.05.2005 - \gtrsymDied~12.05.2023\\
    \bigskip
    This thesis is dedicated to my cat \textbf{Blacky}\\ who almost reached the age of 18 years.\\ She followed me through life since I was 5 years old.\\You always were a big part of my life and\\ 
    \bigskip
    I will miss you.\\
    \bigskip
    Manuel Lippert\\
    \today
\end{center}

### 2023-06-08 bachelorthesis/02-acknowledgement.tex
#39: Correction Cornelia Lippert

First of all, I would like to express my greatest gratitude to my supervisor Arthur Peeters and Florian Rath. I would like to thank both of you for your guidance and support during the submission process of the briefcommunication and the helpful hands during the the progress of this thesis and the restart script. 
\bigskip
I would like to thank Bernhard Winkler and Markus Hilt for the great technical support regarding my questions to the \texttt{btrzx1}-cluster which helps me to develope the restart script as it is now. 
\bigskip
I would also like to thank Paul Schwanitz for his helpfulness in many situations and the good discussions in and outside academia as my best friend. In Addition to that I would like to thank Anna-Maria Pleyer, Dominik Müller and Stefan Barthelmann for the great time sharing the office together. 
\bigskip
I will also thank Anna-Maria Pleyer and my sister Cornelia Lippert for proofreading this thesis and briefcommunication which helps me to get better in spelling.
\bigskip
Outside academia, I would like to express my gratitude to my parents, my brother and his wife, my sister and Hanna Schütz for there encouragement, endurance and support during the time of this thesis.

>>>>

First, I would like to extend my greatest gratitude to my supervisor Prof. Arthur Peeters and Dr. Florian Rath. I would like to thank both of you for your guidance and support during the submission process of the brief communication and the helping hands during the writing process of this thesis and development of the restart script. \bigskip

I would like to thank Bernhard Winkler and Markus Hilt for the great technical support regarding my questions to the \texttt{btrzx1}-cluster which helped me develop the restart script as it is now. \bigskip

Also, I would like to thank my best friend, Paul Schwanitz, for his support in many situations and the good discussions in- and outside of academia. In addition I would like to thank Anna-Maria Pleyer, Dominik Müller and Stefan Barthelmann for the great time of sharing the office together. \bigskip

I thank Anna-Maria Pleyer and my sister Cornelia Lippert for proofreading this thesis and brief communication, which helps me get better at English spelling. \bigskip

Outside of academia, I would like to extend my gratitude to my parents, my brother, his wife and daughter, my sister and my girlfriend for their encouragement, endurance and support while writing this thesis. I am aware that sometimes I am a workaholic who needs to be reminded about the life outside of my academic work.

### 2023-06-08 bachelorthesis/03-abstract.tex
#39: Correction Cornelia Lippert

To reduce the computational effort for the simulations lower input parameter of GKW (Gyro Kinetic Workshop) were tested to find the optimal minimum resolution for the performed simulations. \bigskip

For convenience a \texttt{python} script (\texttt{slurm\_monitor.py}) was written which monitors the simulation on the \texttt{btrzx1}-cluster and starts/restarts until the completion criteria is fulfilled. \bigskip

@@

Ionen-Temperaturgradienten getriebene Turbulenzen (ITG) weisen nahe niedriger Stabilität Zonal FLow Strukturbildungen, sogenannte $\exb$ Treppenstrukturen, auf Mesoskalen auf. Solche Strukturbildungen wurden sowohl in lokal gradientengetriebene Flussschlauchsimulationen als auch in global gradientengetriebenen und global flussgetriebenen Untersuchungen entdeckt. \bigskip

@@

Zur Einfachheit wurde ein \texttt{python}-Skript (\texttt{slurm\_monitor.py}) geschrieben, was Simulationen auf den \texttt{btrzx1}-Cluster überwacht und gegebenenfalls startet/neustartet bis das Kriterium zur Vollendung erfüllt ist. \bigskip

>>>>

To reduce the computational effort for the simulations lower input parameter of GKW (Gyro Kinetic Workshop) were tested to find the optimum of minimum resolution for the performed simulations. \bigskip

For convenience, a \texttt{python} script (\texttt{slurm\_monitor.py}) was written to monitor the simulation on the \texttt{btrzx1}-cluster and start/restart until the completion criterion is fulfilled. \bigskip

@@

Ionen-Temperaturgradienten getriebene Turbulenzen (ITG) weisen nahe niedriger Stabilität Zonal Flow Strukturbildungen, sogenannte $\exb$ Treppenstrukturen, auf Mesoskalen auf. Solche Strukturbildungen wurden sowohl in lokal gradientengetriebene Flussschlauchsimulationen als auch in global gradientengetriebenen und global flussgetriebenen Untersuchungen entdeckt. \bigskip

@@

Der Einfachheit halber wurde ein \texttt{python}-Skript (\texttt{slurm\_monitor.py}) geschrieben, was Simulationen auf den \texttt{btrzx1}-Cluster überwacht und gegebenenfalls startet/neustartet bis das Kriterium zur Vollendung erfüllt ist. \bigskip

### 2023-06-08 bachelorthesis/04-declaration.tex
#39: Correction Cornelia Lippert

The author declares that the material presented in this thesis is his own work, unless explicitly stated otherwise. This thesis is based on
\bigskip
\begin{itemize}
    \item [] \bibentry{Lippert2023}
\end{itemize}
\bigskip
and is a further iteration of this publication (brief communication). It provides additional plots and paragraphs that were not necessary to be included in the final publication. The brief communication can be found in Appendix \ref{append:paper}.

>>>>

The author declares that the materials presented in this thesis are his own work, unless explicitly stated otherwise. This thesis is based on
\bigskip
\begin{itemize}
    \item [] \bibentry{Lippert2023}
\end{itemize}
\bigskip
and is a further iteration of this publication (brief communication). It provides additional plots and paragraphs that were included in the publication. The brief communication can be found in Appendix \ref{append:paper}.

### 2023-06-08 bachelorthesis/10-motivation.tex
#39: Correction Cornelia Lippert

Ion temperature gradient driven turbulence close to marginal stability exhibits zonal flow pattern formation on mesoscales, so-called $\exb$ staircase structures \cite{Pradalier2010}.
Such pattern formation has been observed in local gradient-driven flux-tube simulations \cite{Peeters2016, Weikl2017, Rath2021} as well as global gradient-driven \cite{McMillan2009, Villard2013, Seo2022} and global flux-driven \cite{Pradalier2010, Pradalier2015, Wang2020, Kim2022, Kishimoto2023} studies. 
In global studies, spanning a larger fraction of the minor radius, multiple radial repetitions of staircase structures are usually observed, with a typical pattern size of several ten Larmor radii.
% By contrast, the radial domain size of local flux-tube simulations is often restricted to a similar mesoscale.
By contrast, in the aforementioned local studies the radial size of $\exb$ staircase structures is always found to converge to the radial box size of the flux tube domain.
The above observations lead to the question: \bigskip

@@

In this bachlor thesis the above question is addressed through a box size convergence scan of the same cases close to the nonlinear threshold for turbulence generation as studied in Ref. \citenum{Peeters2016}.
%TODO More Text

>>>>

Ion temperature gradient driven turbulence close to marginal stability exhibits zonal flow pattern formation on mesoscales, so-called $\exb$ staircase structures \cite{Pradalier2010}.
Such pattern formation has been observed in local gradient-driven flux-tube simulations \cite{Peeters2016, Rath2021}, including collisions \cite{Weikl2017} and background $\exb$ shear \cite{Rath2021}, local flux-driven realizations including mean electric field shear \cite{Seiferling2019}, as well as global gradient-driven \cite{McMillan2009, Villard2013, Seo2022} and global flux-driven \cite{Pradalier2010, Pradalier2015, Wang2020, Kim2022, Kishimoto2023} studies. 
In global studies, spanning a larger fraction of the minor radius, multiple radial repetitions of staircase structures are usually observed, with a typical pattern size of several ten Larmor radii.
By contrast, in the aforementioned local studies the radial size of $\exb$ staircase structures is always found to converge to the radial box size of the flux tube domain.
The above observations lead to the question: \bigskip

@@

In this bachlor thesis the above question is addressed through a box size convergence scan of the same cases close to the nonlinear threshold for turbulence generation as studied in Ref. \citenum{Peeters2016}.

### 2023-06-08 bachelorthesis/21-motion.tex
#39: Correction Cornelia Lippert

In magnetic confinement devices like the tokamak reactor, the charged particles experience magnetic and electric fields which results in distinct motion under the associated force. Charged particles can be separated in species, e.g. electrons and ions, which will be later on not displayed in the governing equation. Throughout this thesis the charge $q$ the mass $m$ or the temperature $T$ indicate the quantities of a specific species.

@@

Due to the Lorentz force particles with a velocity component perpendicular to the homogenous magnetic field $v_{\perp}$ undergo a circular motion in the plane perpendicular to the magnetic field. This type motion has circular frequency which is often referred as \textit{cyclotron frequency} and is defined as
\begin{gather}
    \omega_\mathrm{c} = \frac{|q|B}{m}~,
\end{gather}
where $m$ and $q$ are the mass and the charge of the particle and $B$ the strength of the magnetic field. The radius, the so called \textit{Larmor radius}, of this motion is given by
\begin{gather}
    \rho_\mathrm{L} = \frac{mv_{\perp}}{|q|B}
\end{gather}
which center is often referred as \textit{gyrocenter}. Note that since the Lorentz force depends on the species of the particle the circulation direction is opposed between electron in ions.\\\bigskip

Due to the Coulomb collisions the plasma gets thermalized. Together with the Maxwell-Boltzmann distribution the typical thermal velocity is
\begin{gather}
    \vth = \sqrt{\frac{2T}{m}}~,
\end{gather}
where $T$ represents the species temperature. Based on the thermal velocity $v_\mathrm{th}$ the \textit{thermal Larmor radius} gets introduced as
\begin{gather}
    \rhoth = \frac{m\vth}{|q|B}~. \cite{Wesson2011}
\end{gather}

@@

In absence of forces in the parallel direction of the magnetic field the particles can stream freely parallel to the homogenous magnetic field. The velocity of this motion is connected to the thermal velocity $\vth$ and is dominated bei electrons due to their lighter mass compared to ions ($v_\mathrm{th,e}/v_\mathrm{th,i} = 60$).

When an electric field with a component parallel to the magnetic field $E_\parallel$ acts on plasma the charged particles gets accelerated by the electric force
\begin{gather}
    F_{\parallel,E} = qE_\parallel~.
\end{gather}
The parallel motion follows then from the equation of motion. Here the direction of the motion also depends on the species type.\\\bigskip

Since magnetic fields are not always homogenous, an inhomogeneous magnetic field with its gradient $\nabla B$ containing a component parallel to the magnetic field which is given by
\begin{gather}
    \nabla_\parallel B = \frac{\vect{B}}{B} \cdot \nabla B 
\end{gather}
and causes the force
\begin{gather}
    F_{\parallel,\nabla_\parallel B} = - \frac{mv^2_{\perp}}{2B} \nabla_\parallel B = \mu \nabla_\parallel B~;~~~~\mu = \frac{mv^2_{\perp}}{2B}
\end{gather}
with \textit{magnetic moment} $\mu$. The magnetic moment $\mu$ can be an adabatic invariant (constant of motion) if the variation of the magnetic field over time is smaller than the inverse of the cyclotron freqency $\omega^{-1}_\mathrm{c}$ and the spatial variation is larger the Larmor radius $\rho_\mathrm{L}$. The caused force has it application in the mirror effect where a charged particle gets reflected due to this force. \cite{Wesson2011}

@@

\begin{enumerate}
    \item \textbf{$\exb$ Drift:}\\
    If an electric field $\vect{E}$ with a perpendicular component together with the magnetic field $\vect{B}$ (both fields are homogenous) is present the acting Coulomb force and Lorentz force results to a drift of the gyrocenter with
    \begin{gather}
        \vect{v}_{E} = \frac{\vect{E}\times\vect{B}}{B^2}
        \label{eq:exb}
    \end{gather}
    which is called the $\exb$ drift. Since both acting forces direction depends on the species type the direction of the $\exb$ drift is for every species the same.
    \item \textbf{$\nabla B$ Drift:}\\
    Inhomogeneous magnetic field causes a gradient $\nabla B$ of the magnetic field. Because of that gradient the gyrocenter undergoes a $\nabla B$ drift defined by
    \begin{gather}
        \vect{v}_{\nabla B} = \frac{m v^2_{\perp}}{2 q}\frac{\vect{B}\times \nabla B}{B^3}~.
        \label{eq:gradB}
    \end{gather}
    The $\nabla B$ drift varies thereby on scales which are larger compared to the Larmor radius. The direction of the $\nabla B$ drift depends on the species type.
    \item \textbf{Curvature Drift:}\\
    Due to centrifugal force acting on the particle in a curved magnetic filed the gyrocenter experience a curvature drift given by
    \begin{gather}
        \vect{v}_{\kappa} = \frac{m v^2_{\parallel}}{q}\frac{\vect{B}\times \vect{\kappa}}{B^2} = \frac{m v^2_{\parallel}}{q} \frac{\vect{B}\times \nabla B}{B^3}~;\qquad\vect{\kappa} = -(\vect{b}\cdot \nabla)\vect{b} = \frac{\nabla B}{B}~,
        \label{eq:curvature}
    \end{gather}
    where $\vect{b}$ is the unit vector along the magnetic field. To obtain the result for the curvature $\vect{\kappa}$ in Eq. (\ref{eq:curvature}) the plasma pressure has to be small in comparison die the magnetic field strength $B$. In the form of Eq. (\ref{eq:curvature}) $\nabla B$ and curvature drift can be treated similarly. \cite{Wesson2011}
\end{enumerate}

>>>>

In magnetic confinement devices like the tokamak reactor, the charged particles experience forces caused by magnetic and electric fields which results in distinct motion under the associated force. Charged particles can be separated in species, e.g. electrons and ions, which will be later on not displayed in the governing equation. Throughout this thesis the charge $q$, the mass $m$ or the temperature $T$ indicate the quantities of a specific species, i.e., electrons or ions.

@@

Due to the Lorentz force, particles with a velocity component perpendicular to the homogenous magnetic field $v_{\perp}$ undergo a circular motion in the plane perpendicular to the magnetic field. This type of motion has circular frequency, which is often referred to as \textit{cyclotron frequency} and is defined as
\begin{gather}
    \omega_\mathrm{c} = \frac{|q|B}{m}~,
\end{gather}
where $m$ and $q$ are the mass and the charge of the particle and $B$ the strength of the magnetic field. The radius, the so called \textit{Larmor radius}, of this motion is given by
\begin{gather}
    \rho_\mathrm{L} = \frac{mv_{\perp}}{|q|B}
\end{gather}
with the center often being referred to as \textit{gyrocenter}. Note that since the Lorentz force depends on the species of the particle, the circulation direction is the opposite between electron in ions.\\\bigskip

Due to Coulomb collisions the plasma gets thermalized. Together with the Maxwell-Boltzmann distribution the typical thermal velocity is
\begin{gather}
    \vth = \sqrt{\frac{2T}{m}}~,
\end{gather}
where $T$ represents the species temperature. Based on the thermal velocity $v_\mathrm{th}$ the \textit{thermal Larmor radius} gets introduced as
\begin{gather}
    \rhoth = \frac{m\vth}{|q|B}~. \cite{Wesson2011}
\end{gather}

@@

In absence of forces in the parallel direction of the magnetic field the particles can move in parallel direction to the homogenous magnetic field. The velocity of this motion is connected to the thermal velocity $\vth$ and is dominated by electrons due to their lighter mass compared to ions ($v_\mathrm{th,e}/v_\mathrm{th,i} = 60$).

When an electric field with a component parallel to the magnetic field $E_\parallel$ influences the plasma the charged particles are accelerated by the electric force
\begin{gather}
    F_{\parallel,E} = qE_\parallel~.
\end{gather}
The parallel motion follows then from the equation of motion. Here the direction of the motion also depends on the species type.\\\bigskip

Since magnetic fields are not always homogenous, an inhomogeneous magnetic field with its gradient $\nabla B$ containing a component parallel to the magnetic field which is given by
\begin{gather}
    \nabla_{\!\parallel} B = \frac{\vect{B}}{B} \cdot \nabla B 
\end{gather}
causes the force
\begin{gather}
    F_{\parallel,\nabla_{\!\parallel} B} = - \frac{mv^2_{\perp}}{2B} \nabla_{\!\parallel} B = \mu \nabla_{\!\parallel} B~;~~~~\mu = \frac{mv^2_{\perp}}{2B}
\end{gather}
with \textit{magnetic moment} $\mu$. The magnetic moment $\mu$ can be an adabatic invariant (constant of motion) if the variation of the magnetic field over time is smaller than the inverse of the cyclotron freqency $\omega^{-1}_\mathrm{c}$ and the spatial variation is larger the Larmor radius $\rho_\mathrm{L}$. The resulting force has its application in the mirror effect where a charged particle gets reflected due to this force. \cite{Wesson2011}

@@

\begin{enumerate}
    \item {\boldmath $\exb$} \textbf{Drift:}\\
    If an electric field $\vect{E}$ with a perpendicular component together with the magnetic field $\vect{B}$ (both fields are homogenous) is present the acting Coulomb force and Lorentz force results to a drift of the gyrocenter with
    \begin{gather}
        \vect{v}_{E} = \frac{\vect{E}\times\vect{B}}{B^2}
        \label{eq:exb}
    \end{gather}
    which is called the $\exb$ drift. Since both acting forces direction depends on the species type the direction of the $\exb$ drift is for every species the same.
    \item {\boldmath $\nabla B$} \textbf{Drift:}\\
    Inhomogeneous magnetic field causes a gradient $\nabla B$ of the magnetic field. Because of that gradient the gyrocenter undergoes a $\nabla B$ drift defined by
    \begin{gather}
        \vect{v}_{\nabla B} = \frac{m v^2_{\perp}}{2 q}\frac{\vect{B}\times \nabla B}{B^3}~.
        \label{eq:gradB}
    \end{gather}
    The $\nabla B$ drift varies thereby on scales which are larger compared to the Larmor radius. The direction of the $\nabla B$ drift depends on the species type.
    \item \textbf{Curvature Drift:}\\
    Due to centrifugal force acting on the particle in a curved magnetic field the gyrocenter experiences a curvature drift according to
    \begin{gather}
        \vect{v}_{\kappa} = \frac{m v^2_{\parallel}}{q}\frac{\vect{B}\times \vect{\kappa}}{B^2} = \frac{m v^2_{\parallel}}{q} \frac{\vect{B}\times \nabla B}{B^3}~;\qquad\vect{\kappa} = -(\vect{b}\cdot \nabla)\vect{b} = \frac{\nabla B}{B}~,
        \label{eq:curvature}
    \end{gather}
    where $\vect{b}$ is the unit vector along the magnetic field. To obtain the result for the curvature $\vect{\kappa}$ in Eq. (\ref{eq:curvature}) the plasma pressure has to be small compared to the magnetic field strength $B$. In the form of Eq. (\ref{eq:curvature}) $\nabla B$ and curvature drift can be treated similarly. \cite{Wesson2011}
\end{enumerate}

### 2023-06-08 bachelorthesis/22-confinement.tex
#39: Correction Cornelia Lippert

In tokamak devices strong magnetic fields confine the hot plasma. As mentioned in Chapter \ref{sec:motion} a magnetic field forces a perpendicular particle motion and a motion which contains the gyro motion and slow perpendicular gyro center drifts. Because of the much smaller size of the Larmor radius compared to the device size $R$ the particle and energy losses are caused by the gyro center drift. To avoid this at the end of the field lines the magnetic field in the tokamak devices is shaped like a torus. This type of geometry has nested surfaces with constant magnetic flux, so-called \textit{flux sufaces} and with magnetic field lines which appearing on that surfaces. For stability reasons the magnetic field has a toroidal and a poloidal component which also holds equivalence of the plasma pressure. \cite{Stroth2011, Wesson2011} The toroidal component is produced by external coils whereas the poloidal component is provided by the toroidal plasma current. Together the components results in a magnetic field which follows helical trajectories [Fig \ref{fig:confinement}]. To characterize the the quality of confinement the so-called \textit{plasma beta} is used and is given by
\begin{gather}
    \beta = \frac{nT}{\mu_0 B^2/2}~.
\end{gather} 
with $n$ the plasma density, $T$ as temperature, $\mu_0$ the permeability in vacuum and the magnetic field strength $B$. Respectively, the plasma beta compares the thermal plasma pressure $nT$ to the ambient magnetic field pressure $\mu_0 B^2/2$ which indicates when the plasma beta is smaller the confinement is better. In a tokamak reactor the plasma beta has a typical order of few percent. \cite{Wesson2011}
\includegraphicsHere{Theory/Tokamak-Torus.pdf}{
	Toroidal flux surfaces in tokamak plasma with helical magnetic field (\textcolor{Ubtgreen}{green} line) in torus coordinates ($\rho$ (radial), $\phi$ (toroidal), $\theta$ (poloidal)) or zylinder coordinates ($Z$, $R$, $\phi$). \cite{Barton2015}
}{fig:confinement}{0.7}

>>>>

In tokamak devices strong magnetic fields confine the hot plasma. As mentioned in Chapter \ref{sec:motion} a magnetic field forces a perpendicular particle motion and a motion which contains the gyro motion and slow perpendicular gyro center drifts. Because of the much smaller size of the Larmor radius compared to the device size $R$ the particle and energy losses are caused by the gyro center drift. To avoid this at the end of the field lines the magnetic field in the tokamak devices is shaped like a torus. This type of geometry has nested surfaces with constant magnetic flux, so-called \textit{flux surfaces}, and magnetic field lines which appear on these surfaces. To maintain stability the magnetic field has a toroidal and a poloidal component which also is equivalant to the plasma pressure. \cite{Stroth2011, Wesson2011} The toroidal component is produced by external coils whereas the poloidal component is provided by the toroidal plasma current. Together the components result in a magnetic field which follows helical trajectories [Fig \ref{fig:confinement}]. To characterize the quality of confinement the so-called \textit{plasma beta} is used and is given as
\begin{gather}
    \beta = \frac{nT}{\mu_0 B^2/2}~,
\end{gather} 
with $n$ the plasma density, $T$ as temperature, $\mu_0$ the permeability in vacuum and the magnetic field strength $B$. Respectively, the plasma beta compares the thermal plasma pressure $nT$ to the ambient magnetic field pressure $\mu_0 B^2/2$ which indicates that in cases when the plasma beta is smaller the confinement is better. For fusion devices the plasma beta has to be smaller than 1 ($\beta < 1$). In a tokamak reactor the plasma beta has a typical order of a few percent. \cite{Wesson2011}
\includegraphicsHere{Theory/Tokamak-Torus.pdf}{
	Toroidal flux surfaces in tokamak plasma with helical magnetic field (\textcolor{Ubtgreen}{green} line) in torus coordinates ($\rho$ (radial), $\phi$ (toroidal), $\theta$ (poloidal)) or cylindrical coordinates ($Z$, $R$, $\phi$). \cite{Barton2015}
}{fig:confinement}{0.7}

### 2023-06-08 bachelorthesis/23-iontempgrad.tex
#39: Correction Cornelia Lippert

The described electrostatic forces acting on the tokamak plasma causes a variety of collective phenomena. Microinstabilities is one of these phenomena which are instabilities on the spatial scale of the Larmor radius $\rho_\mathrm{L}$ and frequencies much smaller than the cyclotron frequency $\omega_\mathrm{c}$. These instabilities are driven by density and temperature gradients that are present in the tokamak devices. Turbulence caused from microinstabilities are considered as the main reason for the loss of particles energy and momentum from tokamak plasma \cite{Brizard2007, Garbet2010, Horton1999} which limits the confinement time of tokamak reactors. The most dominante instability in tokamak plasma is the ion temperature gradient (ITG) driven instability. \cite{Coppi1967, Cowley1991, Rudakov1961}

@@

The ITG driven instability is present when a sufficiently large ion temperature gradient occurs and is driven by the so-called bad curvature of the outboard low field side of the tokamak. The mechanism will be briefly explained with the help of Fig. \ref{fig:ITG} and Ref. \citenum{Beer_PHD, Casson_PHD, Dannert_PHD}. An ion temperature perturbation ($T^\pm \gtrless 0$ \textcolor{blue}{blue} and \textcolor{red}{red} cells) on the outboard low field side of the tokamak causes the formation of curvature drift and $\nabla B$ dirft with velocity $\vect{v}_\mathrm{d}$ which compresses the plasma ($n_\pm \gtrless 0$, \textcolor{gray}{gray} cells). This relates to the perturbation of the potential ($\phi^\pm  \gtrless 0$) in the adiabatic electron response. As stated in the equation for the electric field ($\vect{E} = -\nabla \phi$) the perturbation of the potential $\phi$ causes an electric field $\vect{E}$ which results in a $\exb$ drift with velocity $\vect{v}_E$. The $\exb$ drifts direction transports hot plasma in the outer hot region of perturbation ($n^+ > 0$, \textcolor{red}{red} cell) and cooler plasma into inner cold region perturbation ($n^- < 0$, \textcolor{blue}{blue} cell) which reinforce the process and instability grows.\bigskip

The described mechanism causes perturbation on the outboard low field side of the tokamak this region has a \enquote{bad curvature} and instability happens there. On the inboard high field side the opposing $\nabla B$ and $\nabla T$ results in a decay of perturbation which is referred to as \enquote{good curvature}. Furthermore, such turbulent structures are called \textit{eddies} which have a typical radial scale of serval Larmor radius. \cite{Newins2006}
Through the explanation above is clear that the initial perturbation is growing when the normalized temperature gradient overcomes a critical value $R/L_\mathrm{T,c}$. For nonlinear turbulence the critical value is given by $R/L_\mathrm{T,c} \sim 6$. \cite{Dimits2000, Isliker2010}

>>>>

The described electrostatic forces acting on the tokamak plasma cause a variety of collective phenomena. The occurence of Microinstabilities is one of these phenomena. They are instabilities on the spatial scale of the Larmor radius $\rho_\mathrm{L}$ and frequencies much smaller than the cyclotron frequency $\omega_\mathrm{c}$, driven by density and temperature gradients present in the tokamak devices. Turbulence by microinstabilities are considered as the main reason for the loss of particle energy and momentum from tokamak plasma \cite{Brizard2007, Garbet2010, Horton1999} which limits the confinement time of tokamak reactors. The most dominant instability in tokamak plasma is the ion temperature gradient (ITG) driven instability. \cite{Coppi1967, Cowley1991, Rudakov1961}

@@

The ITG driven instability is present when a sufficiently large ion temperature gradient occurs and is driven by the so-called bad curvature of the outboard low field side of the tokamak. The mechanism is briefly explained with the help of Fig. \ref{fig:ITG} and Ref. \citenum{Beer_PHD, Casson_PHD, Dannert_PHD}. An ion temperature perturbation ($T^\pm \gtrless 0$ \textcolor{blue}{blue} and \textcolor{red}{red} cells) on the outboard low field side of the tokamak causes the formation of curvature drift and $\nabla B$ drift with velocity $\vect{v}_\mathrm{d}$, which compresses the plasma ($n_\pm \gtrless 0$, \textcolor{gray}{gray} cells). This relates to the perturbation of the potential ($\phi^\pm  \gtrless 0$) in the adiabatic electron response. As stated in the equation for the electric field ($\vect{E} = -\nabla \phi$) the perturbation of the potential $\phi$ causes an electric field $\vect{E}$ which results in a $\exb$ drift with velocity $\vect{v}_E$. The $\exb$ drift direction transports hot plasma in the outer hot region of perturbation ($n^+ > 0$, \textcolor{red}{red} cell) and cooler plasma into inner cold region perturbation ($n^- < 0$, \textcolor{blue}{blue} cell) which reinforce the process and instability grows.\bigskip

The described mechanism causes perturbation on the outboard low field side of the tokamak this region has a \enquote{bad curvature} and instability happens there. On the inboard high field side the opposing $\nabla B$ and $\nabla T$ results in a decay of perturbation which is referred to as \enquote{good curvature}. Furthermore, such turbulent structures are called \textit{eddies} which have a typical radial scale of serval Larmor radius. \cite{Newins2006}
The initial perturbation is growing if the normalized temperature gradient overcomes a critical value $R/L_\mathrm{T,c}$. For nonlinear turbulence the critical value is given by $R/L_\mathrm{T,c} \sim 6$. \cite{Dimits2000, Isliker2010}

### 2023-06-08 bachelorthesis/24-zonalflows.tex
#39: Correction Cornelia Lippert

\section{Zonal Flows and Shearing Rate $\wexb$}
\label{sec:zonalflow}

Zonal flows are linked in the tokamak plasma to the $\exb$ flows [Eq.( \ref{eq:exb})] that is connected to the on flux-surfaces constant electrostatic potential, also called \textit{zonal potential}. \cite{Diamond2005} Because of the variation only across flux-surfaces the zonal flow is tangential to those surfaces and do not contribute to the turbulent transport across the flux-surfaces caused by the $\exb$ motion. 
Zonal flows are driven nonlinear through turbulent Reynolds stresses. \cite{Diamond1991} For more details about the mechanism behind zonal flow generation the reader is referred to Ref. \citenum{Brizard2007, Dubin1983, Hasegawa1978, Krommes2000}. \bigskip

\includegraphicsHere{Theory/Zonal-Flow-Generation.pdf}{
	Subdued turbulence because if stretched eddies which breaks in smaller structures caused by sheared $\exb$ drifts \cite{Hammett2009}
}{fig:zonalflow}{0.9}

Furthermore, zonal flows are connected with the radially alternating zonal potential what introduces a velocity shear, i.e., a radial variation of the zonal flow velocity. This causes a shear deformation of turbulent structures which is considered to play a significant role in the suppression of turbulence \cite{Biglari1990, Dimits2000} especially the ion temperature gradient driven turbulence \cite{Nakata2012, Makwana2014, Maeyama2014, Whelan2018, Whelan2019}. The nonlinear generation mechanism causes a shift, also known as Dimits shift \cite{Dimits2000}, of the critical temperature gradient $R/L_\mathrm{T,c}$ which leads to a later development of ITG driven turbulence. The shearing process tilted und stretches turbulent eddies, caused be the shear flow, until they break into smaller eddies with reduced radial correlation length [Fig. \ref{fig:zonalflow}] which characterize the shearing as decorrelation process. Turbulence will then be reduced because of this process. \cite{Biglari1990,Diamond2005,Burnell1997} \bigskip

@@

When the zonal flows are strong enough to suppress turbulence completely due to the shearing process, pattern formation occurs in the stabilized tokamak plasma the so-called $\exb$ staircase structure. The $\exb$ staircase pattern has the following properties:
\begin{enumerate}
	\item[(1)] They occur on a \textit{radial mesoscale} of order of 10\,$\rhoth$ which is larger than the Larmor radius but smaller than the machine size $R$.
	\item[(2)] The structure are \textit{quasi stationary} in space and varies on time scales much longer than the typical turbulent time scales.
	\item[(3)] They have an \textit{typical amplitude} in terms of the $\exb$ shearing rate of the order of $10^{-1} \vth/R$.
	\item[(4)] Turbulent transport occurs in \textit{avalanches} with propagation strongly linked to the local $\exb$ shearing rate. \cite{McMillan2009}
\end{enumerate} \bigskip

@@

As mentioned before the turbulent transport occurs through avalanches which also mediate the turbulence of the ITG driven turbulence in the adabatic electron approximation. Avalanches are initial when the $\exb$ sharing rate has a zero crossing with a steep flank and propagates outwards with negative shearing and inwards for positive shearing. \cite{Idomura2009,McMillan2009} The fully developed avalanche results in almost completely quenched turbulence. Its development results in a discontinuous step in the dependency of turbulent fluxes on the temperature gradient als known as \textit{finite heat flux threshold}. \cite{Peeters2016,Weikl2017}

>>>>

\section{Zonal Flows and Shearing Rate {\boldmath $\wexb$}}
\label{sec:zonalflow}

Zonal flows are linked in the tokamak plasma to the $\exb$ flows [Eq.(\ref{eq:exb})] that is connected to the on flux-surfaces constant electrostatic potential, also called \textit{zonal potential}. \cite{Diamond2005} Because of the variation exclusively across flux-surfaces, the zonal flow is tangential to those surfaces and does not contribute to the turbulent transport across the flux-surfaces caused by the $\exb$ motion. 
Zonal flows are driven nonlinearly by turbulent Reynolds stresses. \cite{Diamond1991} For more details about the mechanism behind zonal flow generation the reader is referred to Ref. \citenum{Brizard2007, Dubin1983, Hasegawa1978, Krommes2000}. \bigskip

\includegraphicsHere{Theory/Zonal-Flow-Generation.pdf}{
	Subdued turbulence because of stretched eddies, which breaks in smaller structures caused by sheared $\exb$ drifts \cite{Hammett2009}
}{fig:zonalflow}{0.9}

Furthermore, zonal flows are connected with the radially alternating zonal potential, which introduces a velocity shear, i.e., a radial variation of the zonal flow velocity. This causes a shear deformation of turbulent structures which is considered to play a significant role in the suppression of turbulence \cite{Biglari1990, Dimits2000}, especially the ion temperature gradient driven turbulence \cite{Nakata2012, Makwana2014, Maeyama2014, Whelan2018, Whelan2019}. The nonlinear generation mechanism causes a shift, also known as Dimits shift \cite{Dimits2000}, of the critical temperature gradient $R/L_\mathrm{T,c}$ which leads to a later development of ITG driven turbulence. The shearing process tilts and stretches turbulent eddies, caused by the shear flow, until they break into smaller eddies with reduced radial correlation length [Fig. \ref{fig:zonalflow}],  characterizing the shearing as a decorrelation process. Turbulence will then be reduced because of this process. \cite{Biglari1990,Diamond2005,Burnell1997} \bigskip

@@

When the zonal flows are strong enough to suppress turbulence completely due to the shearing process, pattern formation occurs in the stabilized tokamak plasma the so-called $\exb$ staircase structure. The $\exb$ staircase pattern has the following properties:
\begin{enumerate}
	\item[(1)] They occur on a \textit{radial mesoscale} of order of 10\,$\rhoth$ which is larger than the Larmor radius but smaller than the machine size $R$.
	\item[(2)] The structure are \textit{quasi stationary} in space and vary on time scales much longer than the typical turbulence time scales.
	\item[(3)] They have a \textit{typical amplitude} in terms of the $\exb$ shearing rate of the order of $10^{-1} \vth/R$.
	\item[(4)] Turbulent transport occurs in \textit{avalanches} with propagation strongly linked to the local $\exb$ shearing rate. \cite{McMillan2009}
\end{enumerate} \bigskip

@@

As mentioned before the turbulent transport occurs through avalanches which also mediate the turbulence of the ITG driven turbulence in the adabatic electron approximation. Avalanches are initiated when the $\exb$ sharing rate has a zero crossing with a steep flank and propagates outwards with negative shearing and inwards with positive shearing. \cite{Idomura2009,McMillan2009} The fully developed avalanche results in almost completely quenched turbulence. Its development results in a discontinuous step in the dependency of turbulent fluxes on the temperature gradient als known as \textit{finite heat flux threshold}. \cite{Peeters2016,Weikl2017}
